\section{CORE}
\label{sec:exemulcore}

The CORE~\cite{coreemulator} emulator was developed originally by Jeff Ahrenholz at the Boeing Company.
Originally, CORE was an acronym meaning Common Open Research Emulator.
Its official website\footnote{\url{https://www.nrl.navy.mil/itd/ncs/products/core}} is hosted under the Networks and Communication Systems Branch of the U.S. Naval Research Laboratory.

It is a fork of a pre-existing project, Integrated Multiprotocol Network Emulator/Simulator (IMUNES), from the University of Zagreb.

It's described as \textquote{a tool for building virtual networks. As an emulator, CORE builds a representation of a real computer network that runs in real time, as opposed to simulation, where abstract models are used.}
Networks emulated with it can be, via the host machine of the emulation, connected to real world, physical networks.
As its documentation~
\cite{coreghdocs} also states, \textquote{it provides an environment for running real applications and protocols, taking advantage of tools provided by the Linux operating system.}

The first versions of CORE, like its predecessor IMUNES, were based upon both FreeBSD' networking stack and ``jails'' (a particular implementation of containers by that BSD operating system~\cite{freebsdjails})~\cite{comparisonofcore}.

Nowadays, though CORE relies on one of the ways provided by the Linux kernel to enable ``containerization,'' which, as seen earlier in this text, is a way to isolate, at some degree, processes from each other, in this case with special attention to the IO and in particular the OS's networking stack. % TODO acronym for OS

CORE comprises multiple parts, but from a high-level standpoint what matters the most is that it has offers a GUI as the main, official way to interact with the system, and then has a daemon running as a service in the host's Linux setup that is its backend which ensures the creation of namespaces and orchestrates the virtual bridges between them (which support the emulated links).

It is documented to be able to run in a distributed fashion, like GNS3 (cf. ~\ref{sec:exemulgns3}), so that large topologies with a lot of nodes can be put set in motion, distributing that load of virtual nodes across multiple hosts.

The support for a regular layer-3 network with standard routers is provided by Quagga~\cite{quagga}.
All nodes are abstracted as GNU/Linux hosts, which can have a series of \emph{services} running.
The set of services running by default depends on the type (a PC, layer-3 router, layer-2 switch, etc.) that is selected for each node in a project, but can be further customized.
The full list of the services available can be consulted in~\cite{coreemuservices}, but we highlight. % TODO preencher isto e pôr uma tabela

% end of section exemulcore
