% !TEX root = ../Thesis.tex
% !TEX spellcheck = en-US

\chapter{Conclusion}
\label{ch:conclusion}

This chapter offers a summary of the work done throughout this thesis.
It then goes on to highlight ideas that can possibly be useful to explore in the future.

\section{Summary}

This thesis approached the problem of evaluating the suitability of emulators for its use in teaching and learning---how they work, what's necessary to use them, what are their limitations\textellipsis

This work provided a clearer picture of the current situation of emulators.
The technical underpinnings of most of them, which is strictly related to how well they perform in certain cases, and also their weaknesses.
It also clearly shows in what aspects a real course---in particular the one chosen for the case studies, APRC---can benefit from any of the studied emulators starting now.

\section{Future work}

Despite all that was done, there are many more aspects of computer networks studying that were not touch here---some of them are part of the chosen course's curricula (like Open vSwitch and SDN), other are not, like programmable routers with the recent P4 network or NFV.
Seeing how to profit from emulators to introduce new practical topics.

Also, a more fine-grained testing can be done, both in terms of the how these platforms work under scale (more nodes, more traffic between the nodes) but also being used by many students at the time.

Testing the pedagogical results and the methodological fine-tuning necessary in a networks course (say, like APRC) to work exclusively in a software based solution, and measure the degree of success and failure of such an approach is also essential.

Finally, two use cases use-cases were left out of this work that are of interest:
\begin{itemize}
  \item using emulators for application-level exercises (like the ones performed in the FCT/NOVA's Informatics Engineering undergraduate Computer Networks course) and benefit from a reproducible and well-controlled environment;
  \item programming exercises consisting of implementing routing algorithms and protocols, of which the work done in~\cite{teachandlearnmininet} is an example.
\end{itemize}

% end of chapter
